The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-07 kalman_walk_paper.tex
refs and abstract

\begin{abstract}
%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype does not change over evolutionary time, the underlying mechanism may change, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

@@

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model 
% fraisse2016genetics looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations
% chevin2010niche looks at incompats after adaptation in Fisher's geom model and how it depends on dimension
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data

>>>>

\begin{abstract}
%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype does not change over evolutionary time, 
the underlying mechanism may change, 
as distinct molecular pathways can realize identical phenotypes.
Here we use linear system theory to explore the consequences of this idea,
describing how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
This characterization implies that selectively neutral directions in genotype space should be common
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes 
during the early stages of speciation.
\end{abstract}

@@

% wright1932roles mentions possibility of equivalent combinations
% wright1935evolution looks at allele freq dynamics under stabilizing selection on a polygenic quantitative trait and discusses possibility for speciation this way
% barton1989divergence not available but likely following on from barton1986maintenance analyzing multiple optima in a similar model
% nei1983models simulates from simple stepwise mut model on one or two loci (without interactions) where incompat betw diff alleles increases with number of steps betw them
% sved1981twosex mathematically analyzes and (in another paper) simulates speciation in a model of male + female quantitative traits indep'ly evolving with attraction between matching traits
% barton2010natural has a good historical review of natural selection and speciation; then analyzes a two-niche levene model
% gavrilets2004fitness talks a lot about quasineutral paths in holey fitness landscapes
% phillips1996maintenance shows that subpop structure can stably maintain equivalent but incompat genotypes
% tulchinsky2014hybrid simulates a single promoter region with a bioenergentic model with directional and stabilizing selection, found incompat in both cases
% porter2002speciation simulated speciation under a simple model of a regulatory cascade
%%% Fisher's geom model: (incompat due to segregation variance)
% poon2000compensating also uses SSSWM with Fisher's model and derives some analytical results for load in small populations; early citation for this process
% slatkin1994segregation shows that segregation variance increases linearly with divergence if variation is controlled by intermediate freq alleles;
%   but if it is determined by rare alleles (at mut-sel balance), then it plateaus (at twice the within-pop value?)
% barton2001role looks at hybrids betw pops adapting to shifting optima in Fisher's geometric model : under fluctuating optima
%       hybrids will suffer because of cryptic segregating variation
% fraisse2016genetics uses SSWM; looks at stabilizing selection using Fisher's geometric model, showing it predicts lots of observations;
%        but mostly only if fitness is exp(-|x|^k) for k > 2
% chevin2014niche uses SSWM; looks at incompats after adaptation in Fisher's geom model and how it depends on dimension;
%        uses generalized Fisher's model with neutral directions; terms hybrid unfitness due to segregation variance "transgressive incompats";
%        finds rate of RI accumulation decreases with Ne (like Ne^(-m/2), with m the trait dimension?)
% simon2017coadapted follows up on the Fisher's geom model predictions, also comparing to some real data
% martin2014fishers shows that you can get Fisher's model quite generally if (a) the kryptotype dimension is large compared to the phenotype dimension,
%   and (b) mutational effects fall under the CLT
% rosas2010cryptic looks at segregation variance resulting in deviations from optimum phenotype including off of curvey ridges
%%%% General notes
% citation for segregation variance around a stationary optimum -> load is Poon & Otto 2000; rosas2010cryptic
